---
title: "DMV_Phylogenetic_Conservation"
author: "Shruti Kakan"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dependencies: This script relies on the output generated by Rsc04_DMV_size.Rmd, i.e. Combined_data_DMV_with_CpG

```{r}
library(ggplot2)
library(dplyr)
library(scales)
library(geneviewer)
library(RColorBrewer)
library(forcats)
library(tidyr)
library(ggplot2)
library(gtools)
```


### Phylogenetic conservation

#Code to download phastcons 35 way and phylocons 35way bigwig files

```{bash, engine.opts='-l'}
module load system
module load libpng/1.6.53
module load biology
module load ucsc-utils/v489

module load bedtools

REGIONS="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Combined_data_DMV_with_CpG"
OUTDIR="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Phylo_Cons"
mkdir -p $OUTDIR

# ── 1. Download phastCons35way bigWig ────────────────────────────────────────
echo "Downloading phastCons35way bigWig..."
wget -q -O $OUTDIR/phastCons35way.bw \
  "https://hgdownload.soe.ucsc.edu/goldenPath/mm39/phastCons35way/mm39.phastCons35way.bw"

# ── Optional: also download phyloP35way ──────────────────────────────────────
wget -q -O $OUTDIR/phyloP35way.bw \
  "https://hgdownload.soe.ucsc.edu/goldenPath/mm39/phyloP35way/mm39.phyloP35way.bw"

```


Convert raw data for phylogenetic conservation from bigwig to bedGraph prior to intersection. This takes 15 minutes
```{bash, engine.opts='-l'}
module load system
module load libpng/1.6.53
module load biology
module load ucsc-utils/v489

module load bedtools

REGIONS="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Combined_data_DMV_with_CpG"
OUTDIR="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Phylo_Cons"

# ── 2. Convert bigWig to bedGraph ────────────────────────────────────────────
echo "Converting to bedGraph..."
bigWigToBedGraph $OUTDIR/phastCons35way.bw $OUTDIR/phastCons35way.bedGraph
bigWigToBedGraph $OUTDIR/phyloP35way.bw $OUTDIR/phyloP35way.bedGraph


  print
}' $OUTDIR/regions_sorted.bed > $OUTDIR/regions_sorted_chr.bed
```


Sort the combined DMV bed file and add 'chr' before chromosome id, if needed, to match the ucsc phylogenetic conservation files. 
```{bash, engine.opts='-l'}
module load system
module load libpng/1.6.53
module load biology
module load ucsc-utils/v489

module load bedtools

REGIONS="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Combined_data_DMV_with_CpG"
OUTDIR="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Phylo_Cons"

# ── 3. Sort your BED file (bedtools requires sorted input) ───────────────────
sort -k1,1 -k2,2n $REGIONS > $OUTDIR/regions_sorted.bed

# ── 3b. Fix chromosome naming (add chr prefix) ────────────────────────────────
awk 'BEGIN{OFS="\t"} {
  if ($1 !~ /^chr/) $1 = "chr"$1
  print
}' $OUTDIR/regions_sorted.bed > $OUTDIR/regions_sorted_chr.bed
```


```{r}

```

