---
title: "DMV_Phylogenetic_Conservation"
author: "Shruti Kakan"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dependencies: This script relies on the output generated by Rsc04_DMV_size.Rmd, i.e. Combined_data_DMV_with_CpG

```{r}
library(ggplot2)
library(dplyr)
library(scales)
library(geneviewer)
library(RColorBrewer)
library(forcats)
library(tidyr)
library(ggplot2)
library(gtools)
```


### Phylogenetic conservation

#Code to download phastcons 35 way and phylocons 35way bigwig files

```{bash, engine.opts='-l'}
module load system
module load libpng/1.6.53
module load biology
module load ucsc-utils/v489

module load bedtools

REGIONS="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Combined_data_DMV_with_CpG"
OUTDIR="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Phylo_Cons"
mkdir -p $OUTDIR

# ── 1. Download phastCons35way bigWig ────────────────────────────────────────
echo "Downloading phastCons35way bigWig..."
wget -q -O $OUTDIR/phastCons35way.bw \
  "https://hgdownload.soe.ucsc.edu/goldenPath/mm39/phastCons35way/mm39.phastCons35way.bw"

# ── Optional: also download phyloP35way ──────────────────────────────────────
wget -q -O $OUTDIR/phyloP35way.bw \
  "https://hgdownload.soe.ucsc.edu/goldenPath/mm39/phyloP35way/mm39.phyloP35way.bw"

```


Convert raw data for phylogenetic conservation from bigwig to bedGraph prior to intersection. This takes 15 minutes
```{bash, engine.opts='-l'}
module load system
module load libpng/1.6.53
module load biology
module load ucsc-utils/v489

module load bedtools

REGIONS="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Combined_data_DMV_with_CpG"
OUTDIR="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Phylo_Cons"

# ── 2. Convert bigWig to bedGraph ────────────────────────────────────────────
echo "Converting to bedGraph..."
bigWigToBedGraph $OUTDIR/phastCons35way.bw $OUTDIR/phastCons35way.bedGraph
bigWigToBedGraph $OUTDIR/phyloP35way.bw $OUTDIR/phyloP35way.bedGraph


  print
}' $OUTDIR/regions_sorted.bed > $OUTDIR/regions_sorted_chr.bed
```


Sort the combined DMV bed file and add 'chr' before chromosome id, if needed, to match the ucsc phylogenetic conservation files. 
```{bash, engine.opts='-l'}
module load system
module load libpng/1.6.53
module load biology
module load ucsc-utils/v489

module load bedtools

REGIONS="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Combined_data_DMV_with_CpG"
OUTDIR="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Phylo_Cons"

# ── 3. Sort your BED file (bedtools requires sorted input) ───────────────────
sort -k1,1 -k2,2n $REGIONS > $OUTDIR/regions_sorted.bed

# ── 3b. Fix chromosome naming (add chr prefix) ────────────────────────────────
awk 'BEGIN{OFS="\t"} {
  if ($1 !~ /^chr/) $1 = "chr"$1
  print
}' $OUTDIR/regions_sorted.bed > $OUTDIR/regions_sorted_chr.bed
```

#Computing conservation scores for dmv regions, this takes a long time (10 - 15 minutes)
```{bash, engine.opts='-l'}
module load system
module load libpng/1.6.53
module load biology
module load ucsc-utils/v489

module load bedtools

REGIONS="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Combined_data_DMV_with_CpG"
OUTDIR="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Phylo_Cons"

# ── 4.a Extract minimal BED for bedtools intersection ─────────────────────────
# bedtools map -a only needs chr, start, end
cut -f1-3 $OUTDIR/regions_sorted_chr.bed > $OUTDIR/regions_minimal.bed

# ── 4.b Intersect and compute per-region summary stats ────────────────────────
echo "Computing conservation scores..."

#The -null "." flag tells bedtools to write . instead of skipping the row entirely when no overlap is found

# phastCons: mean, max, and fraction of bases covered
bedtools map \
  -a $OUTDIR/regions_minimal.bed \
  -b $OUTDIR/phastCons35way.bedGraph \
  -c 4 -o mean,max,count \
  -null "." \
  > $OUTDIR/phastCons_scores.bed

# phyloP: mean and max (can be negative — accelerated evolution)
bedtools map \
  -a $OUTDIR/regions_minimal.bed \
  -b $OUTDIR/phyloP35way.bedGraph \
  -c 4 -o mean,max \
  -null "." \
  > $OUTDIR/phyloP_scores.bed


# ── 4.c Paste scores back onto original columns ────────────────────────────────
# phastCons: original 7 cols + mean, max, count
paste $OUTDIR/regions_sorted_chr.bed \
      <(cut -f4-6 $OUTDIR/phastCons_scores.bed) \
      > $OUTDIR/regions_phastCons.bed

# phyloP: original 7 cols + mean, max
paste $OUTDIR/regions_sorted_chr.bed \
      <(cut -f4-5 $OUTDIR/phyloP_scores.bed) \
      > $OUTDIR/regions_phyloP.bed

```

#Add headers to the files
```{bash, engine.opts='-l'}
module load system
module load libpng/1.6.53
module load biology
module load ucsc-utils/v489

module load bedtools

REGIONS="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Combined_data_DMV_with_CpG"
OUTDIR="/home/groups/ximenac/01_DMV_Rods/notebooks/All_DMVs/Phylo_Cons"

# ── 5. Add headers ────────────────────────────────────────────────────────────
# For phastCons (expects 10 columns)
awk 'BEGIN{OFS="\t"} NR==1{print; next} {
  while (NF < 10) $(NF+1) = "."
  print
}' $OUTDIR/regions_phastCons.bed > $OUTDIR/regions_phastCons_padded.bed

# For phyloP (expects 9 columns)
awk 'BEGIN{OFS="\t"} NR==1{print; next} {
  while (NF < 9) $(NF+1) = "."
  print
}' $OUTDIR/regions_phyloP.bed > $OUTDIR/regions_phyloP_padded.bed

# Adjust col names to match your BED columns (chr, start, end + any extra)
sed -i '1i chr\tstart\tend\tCpGs\tid\tDMV_type\tDMV_length\tphastCons_mean\tphastCons_max\tphastCons_count' \
  $OUTDIR/regions_phastCons_padded.bed

sed -i '1i chr\tstart\tend\tCpGs\tid\tDMV_type\tDMV_length\tphyloP_mean\tphyloP_max' \
  $OUTDIR/regions_phyloP_padded.bed
  

echo "Done. Output in $OUTDIR"
```


```{r}

```

